% =============================================================================
% PAPER INSERTIONS — statistical significance additions
% =============================================================================
% This file contains drop-in text for each section of the manuscript. Every
% number is from results/per_seed_data.json + results/*.json, computed by
% stats_significance.py. The phrasing follows standard research-paper conventions
% (test name, statistic, dof where applicable, exact p-value, multiple-comparison
% correction noted explicitly).
%
% N = 10 seeds per cell unless otherwise stated. Window for "final cosine" is
% the mean over the last 50 of 300 epochs. Welch's t-test is used for unpaired
% group comparisons; paired t-test is used for within-seed layer comparisons.
% Bonferroni correction is applied to all pairwise families.
% =============================================================================


% -----------------------------------------------------------------------------
% SECTION 2.2 — METRICS  (one sentence at the end of the section)
% -----------------------------------------------------------------------------
% Insert as the final sentence of Section 2.2:

Throughout, group comparisons use Welch's $t$-test, within-seed layer
comparisons use a paired $t$-test, families of pairwise comparisons are
Bonferroni-corrected, and proportion intervals are Clopper--Pearson exact;
all $p$-values are two-sided unless stated otherwise.


% -----------------------------------------------------------------------------
% SECTION 3 — REPLICATION AND THE CHAIN-LENGTH EFFECT
% -----------------------------------------------------------------------------
% Insert immediately AFTER the chain-rule equation (Equation~3) and BEFORE the
% sentence "Each extra layer therefore introduces another mismatched factor...":

A one-way ANOVA on the per-seed final cosine (mean of the last 50 epochs,
$N = 10$ seeds per depth) confirms that depth has a highly significant effect
on alignment, $F(3, 36) = 25.21$, $p = 5.8 \times 10^{-9}$. Bonferroni-corrected
Welch $t$-tests ($m = 6$) show that depth~1 is significantly better aligned
than every deeper network: $t(15.8) = 21.81$, $p_{\mathrm{adj}} < 10^{-11}$
versus depth~2; $t(15.4) = 24.12$, $p_{\mathrm{adj}} < 10^{-11}$ versus
depth~3; and $t(9.1) = 6.22$, $p_{\mathrm{adj}} = 9 \times 10^{-4}$ versus
depth~4. Pairwise comparisons among the deeper networks (2 vs.\ 3, 2 vs.\ 4,
3 vs.\ 4) do not survive correction ($p_{\mathrm{adj}} \in [0.06, 0.13]$),
which reflects the much larger seed-to-seed variance at depth~4
($\mathrm{SD} = 0.16$) caused by the late-training collapse documented in
Section~4. The same one-factor-at-a-time OLS regression as in
Gygax and Zenke (2024) puts confidence intervals on each effect:
the depth coefficient is $-0.106$ ($95\%$~CI $[-0.191, -0.020]$, $p = 0.022$),
the $\log\beta_{\mathrm{sg}}$ coefficient is $+0.069$
($95\%$~CI $[+0.002, +0.135]$, $p = 0.044$), and neither $\log(\text{width})$
nor dataset reaches significance.


% -----------------------------------------------------------------------------
% SECTION 3 — CAPACITY-CONTROLLED DEPTH (after Figure 1 description)
% -----------------------------------------------------------------------------
% Insert immediately AFTER the sentence describing Figure~1 ("...exhibits a
% sharp late-training collapse around epoch 250.") and BEFORE the next
% paragraph "This finding sharpens the original result substantially.":

Pairwise Welch $t$-tests on the capacity-controlled per-seed final cosines
($N = 10$ per depth, Bonferroni $m = 3$) confirm that depth degrades alignment
even when parameter count is held fixed at $\approx\!12{,}737$:
depth~2 (mean $0.803$) vs.\ depth~3 (mean $0.758$) gives
$t(17.9) = 5.07$, $p_{\mathrm{adj}} = 2.4 \times 10^{-4}$, and depth~2 vs.\
depth~4 (mean $0.615$) gives $t(9.3) = 3.68$, $p_{\mathrm{adj}} = 0.014$.
The depth-3-vs-depth-4 comparison is marginal after correction
($t(9.3) = 2.79$, $p_{\mathrm{adj}} = 0.062$), again driven by the elevated
depth-4 variance.


% -----------------------------------------------------------------------------
% SECTION 4 — LATE-TRAINING COLLAPSE
% -----------------------------------------------------------------------------
% Insert immediately AFTER the first sentence stating the collapse rate
% ("...about 30\% of 4-layer moons runs show a sudden, sustained drop..."):

The observed rate of $3/10$ corresponds to a Clopper--Pearson 95\%
confidence interval of $[0.067, 0.652]$; with $N = 10$ the interval is wide
but excludes zero, establishing the existence of the phenomenon while
leaving its true rate uncertain. The mean detected collapse epoch
$\bar{t} = 242.0$ has a $95\%$ $t$-distribution confidence interval of
$[216.8, 267.2]$ (df~$= 2$); this is the conservative reading of the
"$242 \pm 10.1$" reported in the figure caption, which is a standard
deviation rather than a confidence interval.


% -----------------------------------------------------------------------------
% SECTION 4 — SATURATION HYPOTHESIS REJECTION (in the bottom paragraph
% that lists the four failed hypotheses, after "...essentially zero correlation.")
% -----------------------------------------------------------------------------
% Insert as the next sentence after "...Pearson r(saturation, cosine) = 0.008,
% essentially zero correlation.":

A formal test on $N = 3{,}000$ epoch-by-seed pairs confirms the null:
Pearson $r = +0.003$, two-sided $p = 0.875$. We cannot reject the hypothesis
that saturation and cosine alignment are linearly uncorrelated.


% -----------------------------------------------------------------------------
% SECTION 5 — PER-LAYER STORY  (immediately after the gradient-mass values)
% -----------------------------------------------------------------------------
% Insert immediately AFTER "...mean ||g_l||_2 = 0.08, 0.17, 0.15, 0.27 for
% layers 0, 1, 2, readout, the global cosine tracks the readout's cosine
% almost exactly.":

The per-layer separation is statistically very clean. Across $N = 10$ seeds,
the per-seed final cosines (mean of the last 50 epochs) are
$0.056 \pm 0.097$, $0.108 \pm 0.063$, $0.160 \pm 0.061$, and
$0.467 \pm 0.056$ for hidden layers $0$ through $3$ respectively. A
within-seed paired $t$-test (Bonferroni $m = 3$) shows that the readout-side
hidden layer is far better aligned than its neighbour ($t(9) = 10.73$,
$p_{\mathrm{adj}} = 6 \times 10^{-6}$), while the three input-side hidden
layers are statistically indistinguishable from one another (L2 vs.\ L1:
$t(9) = 1.94$, $p_{\mathrm{adj}} = 0.25$; L1 vs.\ L0: $t(9) = 1.17$,
$p_{\mathrm{adj}} = 0.81$). The "input-side layers are perpetually
misaligned" reading is therefore not just a per-seed visual impression
but a tight statistical separation between a single coherent
output-aligned regime and a single coherent input-orthogonal regime.


% -----------------------------------------------------------------------------
% SECTION 6 — beta_sg ROBUSTNESS  (in the existing paragraph, after the
% sentence ending "...drifts from 0.66 to 0.57 as $\beta_{\mathrm{sg}}$
% increases from 2 to 10.")
% -----------------------------------------------------------------------------
% Insert as the next sentence:

The monotone drift in final alignment is itself statistically significant
under an exact two-sided permutation Spearman test
($\rho = -1.0$, $p = 0.017$, $n = 5$ levels), so the small final-cosine
effect of $\beta_{\mathrm{sg}}$ is real even though the collapse rate is
identical at all five steepnesses (a $2 \times 5$ permutation chi-square
gives $p = 1.0$, as expected when every cell is exactly $3/10$).


% -----------------------------------------------------------------------------
% SECTION 6 — POWER RERUN FOR INSTRUMENTATION  (replace or augment the existing
% McNemar sentence with an exact-binomial framing for clarity)
% -----------------------------------------------------------------------------
% Optional: append immediately AFTER "...McNemar p = 0.0654.":

The equivalent exact binomial test on the $11$ discordant pairs
($9$ control-only collapses vs.\ $2$ probed-only collapses) gives
$p = 0.065$ (Newcombe 95\% CI on the rate difference $[-0.060, +0.284]$,
which includes zero). The directional effect is consistent across two
independent statistical lenses but the perturbation result remains
formally non-significant at $\alpha = 0.05$, as already noted in
Section~7.


% -----------------------------------------------------------------------------
% SECTION 2 (METHODOLOGY) — STATISTICAL ANALYSIS SUBSECTION  (optional new
% subsection if the team prefers to consolidate methods in one place)
% -----------------------------------------------------------------------------
% Insert as a new Section~2.4 if desired:

\subsection{Statistical analysis}
\label{sec:stats}
For each cell of the experiment grid we run $N = 10$ random seeds and
report the mean of the cosine similarity over the last $50$ of $300$
training epochs as a per-seed scalar. Group comparisons across
architectures use Welch's $t$-test (unequal variances); within-seed
comparisons across layers use a paired $t$-test. Families of pairwise
tests are Bonferroni-corrected at the family level and reported as
$p_{\mathrm{adj}}$. One-way ANOVA is used to test for any difference
across more than two depths. Exact binomial and Clopper--Pearson
intervals are used for proportions, and paired binary outcomes are
tested with both McNemar's test and the exact binomial on the
discordant pairs. Spearman correlations on small samples (the
$\beta_{\mathrm{sg}}$ sweep, $n = 5$) use exact two-sided permutation
$p$-values. The Welch--Satterthwaite degrees of freedom are reported
to one decimal place. All analyses are reproduced by
\texttt{stats\_significance.py} from the saved per-seed scalars in
\texttt{results/per\_seed\_data.json}.
